Title: A watched pot never boils
Date: 2015-07-13
Author: Duncan Garmonsway
Summary: A mis-optimised kettle
Disqus_identifier: kettle

My tea-crazy flatmate has a fancy kettle, which brings the water to the
temperature you choose.  It is frustratingly slow to boil.

When I suspected that it slowed the rate of heating as it approached the target
temperature, I spent a happy evening heating different volumes of water to
different temperatures to see how the kettle behaved.  Then I lost almost all
the data in a digital equivalent of spilling tea over my lab notes.

Here is the single surviving experiment.

```{r pelican_conf, echo=FALSE}
#set this to true when ready to publish
ready_to_publish = FALSE

# hook_plot <- knit_hooks$get('plot')
# 
# knit_hooks$set(plot=function(x, options) {
#     if (!is.null(options$pelican.publish) && options$pelican.publish) {
#         x <- paste0("{filename}", x)
#     }
#     hook_plot(x, options)
# })
# opts_chunk$set(pelican.publish = ready_to_publish)
knitr::opts_chunk$set(fig.path='./figure/')

knitr::opts_chunk$set(dev = 'Cairo_svg')
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r setup, message = FALSE, warning = FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(scales)
options(lubridate.fasttime = TRUE)
```

```{r}
kettle <- read.csv("./data/kettle.csv", header = TRUE)
```

```{r kettlefig, fig.width = 16, include = TRUE}
kettle %>%
# Remove decreasing steps
mutate(diff = c(NA, diff(Temp))) %>%
filter(diff > 0) %>%
# Start timer at first step
mutate(Time = Time - min(Time)) %>%
# Time in seconds
mutate(Time = origin + seconds(Time)) %>%
# Plot
ggplot(aes(Time, Temp)) +
geom_line() +
scale_x_datetime(name = "Minutes", labels = date_format("%M:%S")) +
scale_y_continuous(name = "Degrees Celsius")
```

Presumably the kettle avoids overshooting, so I tested that in one (lost)
experiment with a very small amount of water and a low target temperature.  It
did, indeed, overshoot, suggesting that the engineers optimised this kettle for
precision, at the expense of speed.

Another flatmate then suggested another experiment to test the reaction speed of
the thermostat, by boiling the kettle, and then chucking a mug of cold water in.
The thermostat responded immediately, which suggests that the problem is in fact
the element.
